---
title: "Plotting Ordinations"
description: Plotting an ordination with base plot and ggplot
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```


This tutorial will show you how to plot an ordination, when you want something more that the default `vegan` ordination plot gives.

We are going to use a data set of aquatic invertebrate abundances in twelve ditches sampled eleven times before and after insecticide treatment. 
The log-transformed species data are available with in the vegan package as `pyrifos`; we have to construct the treatment data from information in the `pyrifos` help file.

```{r library-data}
library("tidyverse")
library("vegan")


#load species data
data(pyrifos)

#create treatment data
pyrifos_env <- tibble(
  week = rep(c(-4, -1, 0.1, 1, 2, 4, 8, 12, 15, 19, 24), each = 12),
  dose = rep(c(0.1, 0, 0, 0.9, 0, 44, 6, 0.1, 44, 0.9, 0, 6), times = 11),
  ditch = rep(letters[1:12], times = 11)
)
```

We can fit a PCA with 

```{r fit-ord}
PCA <- rda(pyrifos)
```

We can make a simple plot of this PCA with the `plot` function.

```{r plot-pca}
plot(PCA)
```

This is a good start, but what if we want to give the observations with different doses of insecticide different colours?

We can get more control over the plotting with `type = "n"` to get an empty plot and then plotting the different components separately.
See `?plot.cca` for more details.

```{r plot-pca2}
plot(PCA, type = "n")
points(PCA, display = "sites", col = factor(pyrifos_env$dose))
```

We could take get a better colour scale or join the observations from the same ditch with a line, but it gets quite complicated.

It can be easier to use `ggplot2` for more complex plots, especially if you are already familiar with `ggplot2`.
This can be done with the `ggvegan` package from [GitHub](https://github.com/gavinsimpson/ggvegan).

The basic plot can be reproduced with `autoplot`.
`autoplot` functions exist for several types of R object.

```{r autoplot}
library("ggvegan")
autoplot(PCA)
```

We can take control of the plotting by using `fortify` to get a data.frame of ordination scores, and then combining this with the treatment data and plotting this with `ggplot`.

```{r fortify}
PCA_fort <- fortify(PCA, display = "sites") # get the scores
head(PCA_fort)

PCA_fort_treat <- bind_cols(PCA_fort, pyrifos_env) %>% 
  mutate(dose = factor(dose),
         dose = fct_rev(dose)) # make dose into a factor and reverse order of levels to improve plot.

PCA_plot <- ggplot(PCA_fort_treat, 
       aes(x = PC1, y = PC2, colour = dose, group = ditch)) +
  geom_point() +
  geom_line() +
  coord_equal() #VERY IMPORTANT or scaling incorrect.

PCA_plot
```

It is important to remember to add the `coord_equal`, or the plot will be incorrectly scales. 

We could add the species to this plot, or make another plot of just the species, perhaps colouring them in some useful way.

This plot could be further improved, but already looks pretty good. 
For example, we could add the variance explained to each axis label, and improve the colour scale.

```{r fortify2}
explained <- eigenvals(PCA)/sum(eigenvals(PCA)) * 100

PCA_plot + #start with first plot
  scale_colour_viridis_d(direction = -1, option = "plasma") + 
  labs(x = glue::glue("PC1 ({explained[1]} %)"), #glue is an alternative to paste for combining text
       y = glue::glue("PC1 ({explained[2]} %)"))

```

## Your turn



