---
title: "Bio303 practical-5 Direct ordination"
description: Atlantic foraminifera
output: html_document
---

```{r setup, include=FALSE}
show_solution = FALSE
knitr::opts_chunk$set(echo = show_solution, eval = show_solution)
```

This practical uses the Imbri and Kipp (1971) dataset of planktic foraminifera relative abundance `atl_spp` and oceanographic conditions `atl_env` at 61 stations in the Atlantic Ocean.


```{r packages}
library("bio303.practicals")
library("vegan")
library("tidyverse")
library("ggvegan")#devtools::install_github("gavinsimpson/ggvegan")
```
Load the data

```{r import}
data("atl_env")
data("atl_spp")
```


## Task 1

Determine if linear or unimodal analyses are likely to be more appropriate?

```{r}
(dca <- decorana(atl_spp))
(dcas <- decorana(sqrt(atl_spp)))  #long. unimodal gradient. complete turnover
```

## Task 2

Select and plot a suitable indirect gradient analysis. Does it exhibit any artefacts? Do other methods behave better?

```{r}
ca <- cca(atl_spp)
plot(ca)
autoplot(ca) # automatic ggplot

#ggplot with more control
fca <- fortify(ca, display = "sites")
head(fca)
fca <- bind_cols(fca, atl_env)

g <- ggplot(fca, aes(x = CA1, y = CA2, colour = WinSST)) + 
  geom_point() +
  coord_equal() +
  scale_colour_gradient(low = "blue", high = "red") + 
  labs(colour = "Winter SST ºC")
g
```

```{r dca}
#plot DCA
autoplot(dca)

fdca <- fortify(dca, display = "sites")
fdca <- bind_cols(fdca, atl_env)

g %+% fdca + aes(x = DCA1, y = DCA2) 
```

```{r pca}
#do alternative methods perform better?
pca <- rda(atl_spp, scale = TRUE)
plot(pca)#horseshoe
```
```{r pcoa}
#metric scaling
pcoa <- capscale(atl_spp ~ 1, distance = "bray")
plot(pcoa, display = "sites")#bad arch

#metric scaling with extended dissimilarities
d <- vegdist(atl_spp, "bray")
plot(dist(atl_env$WinSST), d)#distances saturating at one
d <- stepacross(d, toolong = .9)#use extended similarities
plot(dist(atl_env$WinSST), d)#more linear
pcoa2 <- capscale(d ~ 1, comm = atl_spp)

fpcoa2 <- fortify(pcoa2, display = "sites")
fpcoa2 <- bind_cols(fpcoa2, atl_env)
fpcoa2_spp <- fortify(pcoa2, display = "species") 

g %+% fpcoa2 + 
  aes(x = MDS1, y = MDS2) +
  geom_point(data = fpcoa2_spp, aes(x = MDS1, y = MDS2), inherit.aes = FALSE, shape = 3) +
  ggrepel::geom_text_repel(data = fpcoa2_spp %>% 
  filter(map_dbl(atl_spp, max) > 10), 
            aes(x = MDS1, y = MDS2, label = Label), inherit.aes = FALSE) #labels for common taxa
  
  
#better
```

```{r nmds}
#non-metric scaling
nmds <- metaMDS(atl_spp)
fnmds <- fortify(nmds) %>% 
  filter(Score == "sites")
fnmds <- bind_cols(fnmds, atl_env)

g %+% fnmds + aes(x = NMDS1, y = NMDS2) 

#only slight arch
```
```{r prin-curve}
#principal curve in analogue package
library(analogue)
pc <- prcurve(sqrt(atl_spp), plotit = TRUE)
plot(pc)
atl_env %>% mutate(lambda = pc$lambda) %>% 
  ggplot(aes(WinSST, lambda)) +
  geom_point()
```

```{r best}
#is there an objective test as to which is best? NO. But in an exploratory analysis, we can use 1) test correlation with SST, and 2) check ratio of axis 2 to axis 1 which will give a measure of the strength of the arch in this almost one dimensional dataset 

#correlation of axis one with winter SST
list(dca, ca, pca, nmds, pcoa, pcoa2) %>% 
  map(~scores(.x, choice = 1, disp = "sites")) %>%
  map(~cor(.x, atl_env$WinSST))
#dca, nmds and pcoa with extended dissimilarities have best correlation with SST
cor(pc$lambda, atl_env$WinSST)#highest!

list(ca, pca, pcoa, pcoa2) %>% 
  map(~(.x$CA$eig[2] / .x$CA$eig[1]))
dca$evals[2]/dca$evals[1]
#pcoa with ex diss almost as low as DCA which has guaranteed no arch
#nmds has no eigenvalues. Could devise similar statistics from length of axes.

#Results so far: CA beats PCA, but still does not cope well with this gradient. DCA, pcoa with ex diss and nmds are all OK. Principal curves performs very well

```



## Task 3

Select and plot a suitable direct gradient analysis with all environmental variables? Which variable is more important? What proportion of the variation in the species does it explain?

```{r}
#constrained analysis
#most would use CCA here. Lets start with that.
(cca1 <- cca(atl_spp ~ ., data = atl_env))
plot(cca1)
autoplot(cca1)

fortify(cca1) %>% 
{
    ggplot(filter(., Score == "biplot"), aes(x = CCA1, y = CCA2)) +
    geom_point(data = filter(., Score == "species"), shape = 3) +
    geom_point(
      data = filter(., Score == "sites") %>% 
        bind_cols(atl_env), 
                 mapping = aes(colour = WinSST)) +   
  geom_segment(aes(x = 0, y = 0, xend = CCA1, yend = CCA2), 
               arrow = arrow(length = unit(0.2, "cm")), colour = "navy") +
    geom_text(aes(x = CCA1 * 1.2, y = CCA2 * 1.2, label = Label), colour = "navy") +
    scale_colour_gradient(low = "blue", high = "red") +
    scale_x_continuous(expand = c(0.2, 0)) #make space for labels
  }
  
ca <- cca(atl_spp ~ 1, data = atl_env)#empty model
add1(ca, scope = formula(cca1), test = "perm")#Winter SST is best variable
cca2 <- cca(atl_spp ~ WinSST, data = atl_env)
summary(cca2, display = NULL)#constrained axis accounts for 39.6% of the inertia

```


## Task 4

Partition the variance in the species data between summer temperature and salinity.

```{r}
#can do variance partitioning with varpart. But is only available for RDA, it gives unbiased estimates of the amount explained. (Even a random variable will appear to explain something - varpart uses adjusted-R2 to avoid this problem)
(vp <- varpart(sqrt(atl_spp), atl_env$WinSST, atl_env$Salinity))
plot(vp) #temperature on left, salinity on right.
```

