---
title: "Bio303 - Practical 1"
description: "Introduction to multivatiate methods"
author: "Richard J. Telford"
date: "23rd October 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
```

This practical assumes that you have used R and RStudio before. 
You will probably need to use the help file to complete this practical. 
Type a question mark followed by the name of the function to assess the help system in R. 

## Task 1 - set-up

### Make a course project

Projects help you organise your work in R. 
They help by setting the working directory for you, re-opening your script files when you open a project, and integrating with git and GitHub nicely (see Bio302).

In RStudio, go to `file > new project > new directory` and make a new project. 

You can reopen the project by opening the "project-name.rproj" file, from the file menu, or from the drop-down list in the top right corner of Rstudio.

Load the `bio303.practicals` R package if you haven't already.

```{r load-package}
library("bio303.practicals")
```


### 1.1 Loading data

All the data used in this course are available from the `bio303.practicals` R package to minimise any problems with importing data. 

You can get a list of the available datasets with 

```{r data-list, echo=TRUE, eval = FALSE}
data(package = "bio303.practicals")
```

Datasets can also be loaded with the `data` function

```{r data, echo=TRUE}
data(ponds_spp)
```


The data sets used today are diatom relative abundances and water chemistry data from 20 lakes in England. The data are from (Bennion, H. 1994. A diatom-phosphorus transfer function for shallow, eutrophic ponds in southeast England. _Hydrobiologia_ 275-76: 391-410). 

The species data are in `ponds_spp` and the environmental data in `pond_env`.
Load these data and inspect the objects.


```{r importData}
library("tidyverse")
data("ponds_env")
data("ponds_spp")

head(ponds_env)
head(ponds_spp)
```

The rownames are the pond codes.

### 1.2 The environmental data

Plot a histogram of each environmental variable to see their distribution. 
Do any need transforming to make them more normal?

```{r histograms}
ggplot(ponds_env, aes(x = TP)) +
  geom_histogram(bins = 10)

ponds_env %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 10) +
  facet_wrap(~ variable, scales = "free_x")
```


A simple analysis to find patterns in the environmental data, and check for potential outliers is to run a pairs plot, with plots the variables pairwise. 
You can use `pairs()` or `ggpairs` in the `GGally` package.

```{r pairs}
GGally::ggpairs(ponds_env)
```

### 1.3 Diversity-environment relationships

One way to work with multivariate data is to use them to calculate a univariate summary statistic, such as diversity.

Use function `diversity()` in the vegan library to calculate the Shannon index for each pond.
Use `dplyr::mutate` to add the calculated diversity to the `ponds_env` data.
Plot diversity against total phosphorous (TP) concentration, and use `cor.test()` to test if the correlation is statistically significant.

```{r diversity}
library("vegan")
div <- ponds_env %>% 
  mutate(shannon = diversity(ponds_spp, "shannon"))

ggplot(div, aes(x = TP, y = shannon)) +
  geom_point() +
  geom_smooth(method = "lm")

cor.test(div$TP, div$shannon)
```


### 1.4 Species environment relationships

Choose a common taxon and plot its relative abundance against TP. 
What sorts of relationships do you find? 
Use a GAM (with `gam` from package mgcv) or a GLM with an appropriate error distribution (see bio 302) to determine if these relationships are significant.

```{r speciesEnvironment}
library("mgcv")

(common <- which(colSums(ponds_spp > 0) > 15))

tibble(TP = ponds_env$TP, AC001A = ponds_spp$AC001A/100) %>% 
  ggplot(aes(x = TP, y = AC001A)) +
  geom_point() +
  geom_smooth(method = "gam", method.args = list(family = "quasibinomial"))

mod <- gam(ponds_spp$AC001A/100 ~ s(TP), data = ponds_env, family = quasibinomial, weights = rep(100, nrow(ponds_spp)))
anova(mod)

ponds_spp %>% 
  select(any_of(names(common))) %>% 
  mutate(TP = ponds_env$TP) %>% 
  pivot_longer(cols =  -TP, names_to = "species", values_to = "abundance") %>% 
  mutate(abundance = abundance/100) %>% 
  ggplot(aes(x = TP, y = abundance, colour = species)) +
  geom_point() +
  geom_smooth(method = "gam", method.args = list(family = "quasibinomial"), se = FALSE)

```

### 1.5 Distance metrics

Use `vegdist()` in the vegan package to calculate the Bray-Curtis distance between sites in terms of species composition. 
Calculate the correlation between this and the Euclidean distance in TP space, and plot the two distance matrices against each other. 
Use a `mantel()` test to determine if the correlation is significant?

```{r euclid}
dTP <- dist(ponds_env$TP)
dbray <- vegdist(ponds_spp, "bray")

tibble(dTP = as.vector(dTP), dbray = as.vector(dbray)) %>% 
  ggplot(aes(x = dTP, y = dbray)) +
  geom_point() +
  geom_smooth()

cor(dTP, dbray) 
mantel(dTP, dbray) #cannot use cor.test as number of degrees of freedom is wrong
```

How does the correlation change when all (scaled) variables are included in the distance calculation? 
What does that tell you about the ecological importance of at least some of the other variables?

```{r scaledEuclid}
denv <- dist(scale(ponds_env))

tibble(denv = as.vector(denv), dbray = as.vector(dbray)) %>% 
  ggplot(aes(x = denv, y = dbray)) +
  geom_point() +
  geom_smooth()

cor(denv, dbray)
mantel(denv, dbray)
```

Use `bioenv()` to select the best subset of variables (allow up to 5 variables or the analysis will take ages). Interpret the result.

```{r bioenv}
bioenv(ponds_spp, ponds_env, upto = 5)
```

Repeat this analysis using Euclidean distances for the species data. 
How do the correlations change? 
What does this imply?

```{r bioenv_Euclid}
deuc <- dist(ponds_spp)
mantel(dTP, deuc)
bioenv(ponds_spp, ponds_env, upto = 5, index = "euclidean")
```




