---
title: "Bio303 practical-4 Direct ordination"
description: Adding the environmental data
output: html_document
---

```{r setup, include=FALSE}
show_solution = FALSE
knitr::opts_chunk$set(echo = show_solution, eval = show_solution)
```



In this practical we will attempt to find which environmental variables are controlling diatom assemblages in English ponds using indirect and direct gradient analysis.

## Task 1

Load the ponds species and environmental data, and necessary packages.

```{r load-data}
library("bio303.practicals")
library("tidyverse")
library("vegan")
data("ponds_env")
data("ponds_spp")
```


## Task 2

Run a DCA analysis of the species data using `decorana()` and determine if a linear or unimodal model is most appropriate. It may be useful to use a `sqrt` transformation to stabilise the variances.

```{r decorana}
#axis length
decorana(ponds_spp)
decorana(sqrt(ponds_spp))

#large reduction in first axis length with square root transform. Probably noise.
```


## Task 3

Use indirect gradient analysis to ordinate the species data. There are several sensible choices for methods: 

- PCA of `sqrt` transformed data with function `rda` (equivalent to using chord distance)
- CA with function `cca` with either raw or `sqrt` transformed data
- PCoA on Bray-Curtis distances using function `capscale` (this function requires a formula – use `ponds_spp ~ 1` for unconstrained ordination).
- NMDS with function `metaMDS` (you need to choose how many axes to recover)

Plot a `screeplot()` of the ordination with argument `bstick = TRUE` (not for PCoA or NMDS). Determine how many axes are interpretable. Plot the ordination. The ordination plots are fully adjustable, see the example for `?plot.cca()` and the plotting tutorial

```{r indirect}
#indirect ordination
ord <- rda(decostand(ponds_spp, "hellinger"))
screeplot(ord, bstick = TRUE)#two significant axes


par(mfrow = c(2, 2))
screeplot(cca(ponds_spp), bstick = TRUE)
screeplot(cca(sqrt(ponds_spp)), bstick = TRUE) #sqrt improves noisy data
screeplot(rda(ponds_spp), bstick = TRUE)
screeplot(rda(sqrt(ponds_spp)), bstick = TRUE)
```

```{r}
plot(ord)#default plot - horseshoe. Change to CA

ord <- cca(sqrt(ponds_spp))
plot(ord, type = "n") #empty plot
points(ord, disp = "species", pch = "+", col = 2) #add components to empty plot
points(ord, disp = "sites", pch = 20)
text(ord, disp = "species", select = colSums(ponds_spp > 0) < 5, cex = .5, col = 2)
#couple sites extreme on axis 2 might be driven by rare taxa
```

```{r ggvegan}
library("ggvegan")#installed from github

#remove taxa with 2 or fewer occurrences
ord <- cca(sqrt(select_if(ponds_spp, colSums(ponds_spp > 0) > 2)))

#default plot
autoplot(ord)

#with full control
ord_f <- fortify(ord)#make into a data_frame
head(ord_f)
ord_f %>% filter(Score == "sites") %>% 
  ggplot(aes(x = CA1, y = CA2)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_hline(yintercept = 0, colour = "grey60") + 
  geom_point() +
  coord_equal() + #don't forget!!
  labs(
    x = paste0("PCA1 (",round(eigenvals(ord)[1]/ord$tot.chi * 100, 1),"%)"),
    y = paste0("PCA2 (",round(eigenvals(ord)[2]/ord$tot.chi * 100, 1),"%)")
    )

```

## Task 4

Determine which variables are significantly correlated with the first two ordination axes with `envfit()`, using 999 permutations. Use the envfit’s plot function to add the significant variables to the ordination plot as arrows.

```{r envfit}
(efit <- envfit(ord, ponds_env, perm = 1000))
plot(ord)#baseplot
plot(efit)#everything
plot(efit, p.max = .05, col = "green")#just significant ones
```

## Task 5

It can be more useful to show environmental variables as surfaces rather than arrows, as they are often not linearly correlated with ordination space. This can be done with function `ordisurf()`, with uses a GAM to smooth and interpolate the environment. Do this for the environmental variables found to be significant by `envfit`.


```{r}
#ordisurf
ordisurf(ord, ponds_env$Chla)
```

## Task 6

Run an appropriate direct gradient analysis using all environmental variables. Use the formula notation `ponds_spp ~ .` and the argument `data = ponds_env`. The “.” means include all variable in the model.

- RDA of `sqrt` transformed data with function `rda` (equivalent to using chord distance)
- CCA with function `cca` with either raw or `sqrt` transformed data
- dRDA on Bray-Curtis distances using function `capscale` or `dbrda`

How much of the variation is explained by the first axis? How does this compare with the unconstrained ordination? Plot the analysis.

```{r}
mod <- rda(decostand(ponds_spp, "hellinger") ~ ., data = ponds_env)#all variables
mod
ord
#summary(mod) # lots of information
plot(mod)
```


## Task 7

Many of the variables are not significant, or are collinear (use `vif.cca` to find the variance inflation factors), and should be removed to improve the interpretability of the results. Use forward selection with function `ordistep` to determine a good subset of variable that gives a more parsimonious model.

```{r}
vif.cca(mod)#some high values - indicative of problem

#automatic selection 
minmod <- step(mod, test = "perm")#starting with the full model
vif.cca(minmod)#all now low

minmod2 <- step(rda(decostand(ponds_spp, "hellinger") ~  1, ponds_env), reformulate(names(ponds_env)), test = "perm")#starting with an empty model - gives more parsimonious solution

#manual model building
## -- define the maximal model for scope
mbig <- rda(decostand(ponds_spp, "hellinger") ~  ., ponds_env)
## -- define an empty model to start with
m0 <- rda(decostand(ponds_spp, "hellinger") ~ 1, ponds_env)
## -- manual selection and updating
add1(m0, scope = formula(mbig), test = "perm")#start with empty model - test which variable is best to add
m0 <- update(m0, . ~ . + TP)#tp gives best (lowest) AIC and is significant
add1(m0, scope = formula(mbig), test = "perm")
m0 <- update(m0, . ~ . + Alkalinity)
## -- included variables still significant?
drop1(m0, test = "perm")
add1(m0, scope = formula(mbig), test = "perm")
m0 <- update(m0, . ~ . + Secchi)
drop1(m0, test = "perm")
add1(m0, scope = formula(mbig), test = "perm")
m0 <- update(m0, . ~ . + Maxdept)
drop1(m0, test = "perm")
add1(m0, scope = formula(mbig), test = "perm")#no more significant variables
plot(m0)

#different ordinations methods may find different sets of variables to be important
```

