---
title: "3 - Indirect ordination"
description: "How well do indirect ordinations work?"
author: "Richard J. Telford"
date: "23rd October 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
```

The aim of indirect gradient analysis is to try to recover the latent structure in the data. 
It is impossible to assess how successful this recovery is with real data as the true structure is not known. 
This makes tests with simulated data, where the structure is know, useful. 

In this practical, you will try to recover the structure in two simulated datasets, and compare the recovered structure to the true structure to determine which ordination procedures work well. 
The data are simulated using a procedure similar to that developed in Minchin’s (1987) COMPAS program or Gavin Simpson's [`coenocliner` package](https://github.com/gavinsimpson/coenocliner).

Both data sets have has 100 sites arranged in a 2-dimensional grid and about 30 species. 
In dataset 1  (`sim_spp1` and `sim_env1`), both environmental gradients are the same length, in dataset 2 (`sim_spp2` and `sim_env2`), the second gradient is about three times as long as the first.
The species data are simulated abundances.

## Task 1.1

Load the `bio303.practicals` package if you have not already done so.

```{r load, echo = TRUE}
library("bio303.practicals")
```


Import and inspect the simulated data for the first dataset.

```{r load-sim1, echo = TRUE}
data(sim_env1)
data(sim_spp1)

head(sim_env1)
head(sim_spp1)
```


## Task 1.2

Use the `plot_grid` function to show the shape of the environmental data in dataset 1.
In the plot, there is a site at each node of the grid, and the lines join sites with the same values of the two environmental variables.
Coloured lines mark the middle of each environmental variable.

```{r plot-env1-grid, echo = TRUE}
plot_grid(sim_env1)
```


## Task 1.3

Now load the `vegan` package which we will use for all ordinations.

```{r load-vegan, echo = TRUE}
library("vegan")
```


Run function `decorana()` from the vegan package on dataset `sim_spp1`. 
This will fit a Detrended Correspondence Analysis to the species data.

- Examine the output. How long is the first axis? 
- What type of analysis is appropriate – one that assumes linear or unimodal responses?
- Plot the DCA 

```{r decorana, echo = TRUE}
#length first axis
dca1 <- decorana(sim_spp1)
dca1

#dca
plot(dca1)
```


## Task 1.4

It is see how the DCA has distorted the actual gradients.
This is where we can use `plot_grid` to highlight the distortions in the original grid.


```{r plot-grid-dca1, echo = TRUE}
plot_grid(dca1)#note triangle configuration

plot_grid(dca1, choice = 2:3)#note twist in axis 3 - very ugly
```


## Task 1.5

We can quantify our comparison of the original and retrieved structure with a Procrustes rotation. 

Proctustes rotations mirror, rotate and stretch ordinations so that they match as well as possible and the residual differences can be quantified.

Run code like `procrustes(sim_env1, DCA1, symmetric = TRUE)`, where `DCA1` is an ordination object.
The Procrustes sum of squares is the measure of success, being zero for a perfect match.

```{r procrustes, echo = TRUE}
pro <- procrustes(sim_env1, dca1, symmetric = TRUE, choices = 1:2)
pro
plot(pro)
```

## Task 1.6

Repeat tasks 1.3 to 1.5 with different ordination methods.
Try to find the method that gives the lowest Procrustes sum of squares. 

Try:

-	PCA with function `rda`
-	CA with function `cca`
-	dbRDA with function `dbrda`
-	NMDS with function `metaMDS`


```{r}
#pca
pca1 <- rda(sim_spp1)
plot_grid(pca1)
procrustes(sim_env1, pca1, symmetric=TRUE)

pca1s <- rda(decostand(sim_spp1, "hellinger"))
plot_grid(pca1s)
procrustes(sim_env1, pca1s, symmetric = TRUE) #good result


#ca
ca1 <- cca(sim_spp1)
plot_grid(ca1)
procrustes(sim_env1, ca1, symmetric = TRUE)

ca1s <- cca(decostand(sim_spp1, "hellinger"))
plot_grid(ca1s)
procrustes(sim_env1, ca1s, symmetric = TRUE)


#dbrda
dbrda1 <- dbrda(sim_spp1 ~ 1, distance = "bray")
plot_grid(dbrda1)
procrustes(sim_env1, dbrda1, symmetric = TRUE)

#dbrda with NMDS data processing
dbrda2 <- dbrda(sim_spp1 ~ 1, distance = "bray", metaMDSdist = TRUE)
plot_grid(dbrda2)
procrustes(sim_env1, dbrda2, symmetric = TRUE)

#nmds 
nmds1 <- metaMDS(sim_spp1, k = 2)
plot_grid(nmds1)
procrustes(sim_env1, nmds1, symmetric = TRUE)

```


Other things to try:

-	Various data transformations with function decostand
-	Different treatments for rare taxa 

## Task 2

Repeat the above with the second dataset which has a different shape.
Does the same method give the best result?


```{r}
#dataset 2

plot_grid(sim_env2)

#length first axis
(dca2 <- decorana(sim_spp2))

#dca
plot_grid(dca2)
procrustes(sim_env2, dca2, symmetric = TRUE, choice = 1:2)

#pca
pca2 <- rda(sim_spp2)
plot_grid(pca2)
procrustes(sim_env2, pca2, symmetric = TRUE)

pca2 <- rda(decostand(sim_spp2, "hellinger"))
plot_grid(pca2)
procrustes(sim_env2, pca2, symmetric = TRUE)


#ca
ca2 <- cca(sim_spp2)
plot_grid(ca2)
procrustes(sim_env2, ca2, symmetric = TRUE)

#dbrda
dbrda2 <- dbrda(sim_spp2~1, distance="bray")
plot_grid(dbrda2)
procrustes(sim_env2, dbrda2, symmetric = TRUE)

#pcoa 
dbrda22 <- dbrda(sim_spp2~1, distance="bray", metaMDSdist = TRUE)
plot_grid(dbrda22)
procrustes(sim_env2, dbrda22, symmetric = TRUE)

#nmds 
nmds2 <- metaMDS(sim_spp2, k = 2)
plot_grid(nmds2)
procrustes(sim_env2, nmds2, symmetric = TRUE)
```

Remember that these simulated data have relatively little noise, and symmetrical responses. The methods also need to be tested with more challenging simulations.